Estimating duration using survival analysis

### 2023-05-30 cis-perfect-testing.tex
Make i.e. prettier

This chapter aims to estimate the duration of SARS-CoV-2 PCR positivity from CIS data if perfect testing can be assumed (i.e.: no misclassification bias).
I take a survival analysis approach, considering that neither the start nor end time of positivity are known exactly (double interval censoring) and that infections may be missed (truncation), explained in the following paragraphs.
Through a simulation study, I show the applicability of these methods.
However, application in reality must account for misclassification bias, the subject of the following chapter.

@@

I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \autoref{imperf-test}).
The episode's \emph{duration of positivity} is the number of days that they are detectable for; I use time discretised to days and hence this is at least 1 day.
Note that the relationship between positivity and infectiousness is complex\todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).

Series of positive tests within CIS were heuristically categorised into episodes by Sarah Walker (see \autoref{episode-def}).
These tests are assumed to have been due to a single infection event.

I take a survival analysis approach to analysing this data.
Survival analysis is the area of statistics which estimate the distribution of times between events.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In this context, the distribution to estimate is that of duration of positivity.
The \emph{initiating event} is when the duration starts, here becoming detectable.
The \emph{terminating event} is when the duration starts, here becoming no longer detectable.

@@

The first issue is double interval censoring.
An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
Censoring is a very widely studied problem, and a central consideration in any survival analysis.
Interval censoring arises in CIS because we only observe the change from being not detectable (i.e.: negative) at one test, to detectable (i.e.: positive) at the following test and vice versa; \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring (see \autoref{perf-test:fig:double-interval-censor}).
\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored.
\begin{figure}
  \todo[inline]{Copy over figure demonstrating double interval censoring}
  \caption{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. \label{perf-test:fig:double-interval-censor}}
\end{figure}

@@

I use the CIS data described in section\todo{ref description of CIS data}, with positive tests heuristically divided into episodes.
The episode definition was developed by Sarah Walker and is as follows~\citePersonalComms:
\todo[inline]{Insert episode definition}

I filter the episodes as follows.
\begin{itemize}
\item
  The first positive in the episode occurred between 16th Oct 2020 and
  5th Dec 2020 inclusive.
\item
  The individual had recorded a negative test prior to their episode
  beginning (this could have happened at any time, including prior to
  16th Oct).
\item
  A negative test to end the episode (i.e.: following the final positive in the episode) has been recorded.
\end{itemize}

@@

The individual model considers only individuals with observed (i.e., not truncated) episodes.
This approach is similar to methods used in previous studies, such as \textcite{heiseyModelling}, \textcite{dempsterMaximum}, and \textcite{turnbullEmpirical}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
The modelling approach then corrects for this selection bias.

@@

Formally (and following \textcite{heiseyModelling}), we
define three sets which partition the space of possible infection and
recovery (i.e.: are strict subsets of $T \times T$) given the
observations associated with episode $i$ (graphically shown in Figure
\ref{perf-test:fig:partitionSpace}). Each episode, whether observed or not, must
fall into one of these three classes with a probability that is a
function of $\theta$ (see section \ref{perf-test:sec:likelihood}).

@@

The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
Outside this period, all tests are negative (i.e.: no false positives occur). 
Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).

>>>>

This chapter aims to estimate the duration of SARS-CoV-2 PCR positivity from CIS data if perfect testing can be assumed (\ie: no misclassification bias).
I take a survival analysis approach, considering that neither the start nor end time of positivity are known exactly (double interval censoring) and that infections may be missed (truncation), explained in the following paragraphs.
Through a simulation study, I show the applicability of these methods.
However, application in reality must account for misclassification bias, the subject of the following chapter.

@@

I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \autoref{imperf-test}).
The episode's \emph{duration of positivity} is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
Note that the relationship between positivity and infectiousness is complex, \todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.

An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).

I utilise survival analysis for analysing this data.
Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In the context of this thesis, the distribution to estimate is that of duration of positivity.
The \emph{initiating event} is when the duration starts, \ie becoming detectable.
The \emph{terminating event} is when the duration starts, \ie becoming no longer detectable.

@@

The first issue is double interval censoring.
An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
Censoring is a very widely studied problem, and a central consideration in any survival analysis.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa; \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring (see \autoref{perf-test:fig:double-interval-censor}).
\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored.
\begin{figure}
  \todo[inline]{Copy over figure demonstrating double interval censoring}
  \caption{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. \label{perf-test:fig:double-interval-censor}}
\end{figure}

@@

I use the CIS data described in section\todo{ref description of CIS data}, with positive tests heuristically divided into episodes.
Prior work categorised a series of positive tests within the same individual CIS were heuristically categorised into episodes (see \autoref{episode-def}).

I filter the episodes as follows.
\begin{itemize}
\item
  The first positive in the episode occurred between 16th Oct 2020 and
  5th Dec 2020 inclusive.
\item
  The individual had recorded a negative test prior to their episode
  beginning (this could have happened at any time, including prior to
  16th Oct).
\item
  A negative test to end the episode (\ie: following the final positive in the episode) has been recorded.
\end{itemize}

@@

The individual model considers only individuals with observed (\ie, not truncated) episodes.
This approach is similar to methods used in previous studies, such as \textcite{heiseyModelling}, \textcite{dempsterMaximum}, and \textcite{turnbullEmpirical}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
The modelling approach then corrects for this selection bias.

@@

Formally (and following \textcite{heiseyModelling}), we
define three sets which partition the space of possible infection and
recovery (\ie: are strict subsets of $T \times T$) given the
observations associated with episode $i$ (graphically shown in Figure
\ref{perf-test:fig:partitionSpace}). Each episode, whether observed or not, must
fall into one of these three classes with a probability that is a
function of $\theta$ (see section \ref{perf-test:sec:likelihood}).

@@

The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
Outside this period, all tests are negative (\ie: no false positives occur). 
Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).
